---
title: "data_wrangle"
output: html_document
date: "2024-12-06"
---

```{r setup, include=FALSE}
if(!require("pacman"))
  install.packages("pacman")

pacman::p_load(
  # need here for loading data to all systems
  here,
  dplyr,
  tidyr,
  usmap,
  janitor
)
```

```{r load}
data_cancer <- read.table(here("data", "cancer.txt"), header = TRUE)
data_race <- read.table(here("data", "race.txt"), header = TRUE)
# data_race_supp <- read.table(here("data", "race_supp.txt"), header = TRUE)
data_sunlight <- read.table(here("data", "sunlight.txt"), header = TRUE)
```

`data_race_supp` is because the census has adjusted its categorization, so 2021 was missing from `data_race`.
_Update: sunlight data only goes to 2011 so this didn't matter..._

```{r cleanup}
clean_data <- function(input_data){
  na_vals <- c("Missing", "Not Applicable", "Not Available")
  int_cols <- c("states_code", "year")
  numeric_cols <- c("count", "population", "age_adjusted_rate", "crude_rate")
  rename_cols <- c(
    "state_code" = "states_code",
    "state" = "states",
    "year" = "yearly_july_1st_estimates",
    "year_code" = "yearly_july_1st_estimates_code"
    )
  
  cleaned_data <- input_data |> 
    clean_names() |> 
    rename(any_of(rename_cols)) |> 
    select(!any_of("year_code")) |> 
    mutate(
      across(everything(), ~replace(., . %in% na_vals, NA)),
      across(any_of(int_cols), ~as.integer(.)),
      across(any_of(numeric_cols), ~as.numeric(.))
    )
  
  return(cleaned_data)
}

data_cancer <- clean_data(data_cancer) |> select(!population)
data_race <- clean_data(data_race)
# data_race_supp <- clean_data(data_race_supp) |> filter(race_code != 999)
data_sunlight <- clean_data(data_sunlight)
```


```{r index}
# data_race_w <- rbind(data_race, data_race_supp) |> 
data_race_w <- data_race |> 
  mutate(race_w = ifelse(race == "White", "white", "non_white")) |> 
  group_by(state_code, year, race_w) |>
  summarize(population = sum(population, na.rm = TRUE)) |> 
  pivot_wider(values_from = population, names_from = race_w) |> 
  mutate(
    total_pop = non_white + white,
    prop_white = white/total_pop
  )
```

Uniting the data.

Some states are `Missing` cancer count data for some years (not just 0), so they will be dropped.
Note that instances of a cancer count of 0 will have an `NA` for `cancer_age_adj_rate` and `cancer_crude_rate`.

```{r combine}
data_full <- data_cancer |> 
  drop_na(count) |> 
  inner_join(data_race_w, by = join_by(year, state_code)) |> 
  inner_join(select(data_sunlight, !state), by = join_by(year, state_code)) |> 
  arrange(year, state_code) |> 
  select(
    year,
    fips = state_code,
    state,
    cancer_count = count,
    proportion_white = prop_white,
    sunlight_daily_avg = avg_daily_sunlight_kj_m2,
    cancer_age_adj_rate = age_adjusted_rate,
    cancer_crude_rate = crude_rate,
    population_total = total_pop,
    population_non_white = non_white,
    population_white = white,
    everything()
  ) |> 
  mutate(fips = sprintf("%02s", as.character(fips)))
```

I used the FIPS code (`fips`) throughout, but using the regular state names (`full` in `data_us` shape file, `state` in others) should also work fine when uniting data.

Not all states have sunlight data, so I removed those from the map.

D.C. (and AK, which also doesn't have sunlight data) have a greater population of non-White folks, so if this one index number is messing with the analysis we could consider dropping it.

```{r shape_files}
include_states <- unique(data_full$fips)

data_us <- us_map(include = c(as.character(include_states)))
```

```{r save_data}
saveRDS(data_full, file = here("data", "data_full.rds"))
```

