---
title: "data_analysis"
output: html_document
date: "2024-12-08"
---


```{r setup, include=FALSE}
if(!require("pacman"))
  install.packages("pacman")

pacman::p_load(
  here,   # need here for loading data to all systems
  dplyr,
  tidyr,
  usmap,
  ggplot2, 
  leaflet,
  sf,
  spdep, #for computing spatial objects to for Moran's I
  CARBayesST

)
```

```{r load data}
#load data from wrangling
data <- readRDS(here("data","data_full.rds"))
data <- filter(data, year>=2003)

#open US shape file
include_states <- unique(data$fips)
data_us <- us_map(include = c(as.character(include_states)))
data_us <- st_transform(data_us, crs = 4326) #convert to lat long
st_crs(data_us) #confirm we are in lat long

```

```{r eda}

#EDA of covariates through the years

#--------
#sunlight
#--------

ggplot(data, aes(x = factor(year), y = sunlight_daily_avg)) +
    geom_boxplot(fill="red", alpha=0.7) + 
    labs(title = "Boxplot of daily average sunlight") + 
    scale_x_discrete(name = "Year", 
                     breaks=c(2003, 2004, 2005, 2006, 2007 ,2008, 2009, 2010, 2011), 
                     labels=c("2003", "2004", "2005", "2006", "2007" , "2008", "2009", "2010", "2011")) +
    scale_y_continuous(name = expression(paste("Daily Sunlight Average (W/m"^"2)"))) + 
    theme(text=element_text(size=16), plot.title=element_text(size=18, face="bold"))

#--------
#cancer count
#--------


ggplot(data, aes(x = factor(year), y = cancer_count)) +
    geom_boxplot(fill="blue", alpha=0.7) + 
    labs(title = "Boxplot of cancer count") + 
    scale_x_discrete(name = "Year", 
                     breaks=c(2003, 2004, 2005, 2006, 2007 ,2008, 2009, 2010, 2011), 
                     labels=c("2003", "2004", "2005", "2006", "2007" , "2008", "2009", "2010", "2011")) +
    scale_y_continuous(name = expression(paste("Cancer count per 10,000 people"))) + 
    theme(text=element_text(size=16), plot.title=element_text(size=18, face="bold"))

#--------
#proportion white
#--------

ggplot(data, aes(x = factor(year), y = proportion_white)) +
    geom_boxplot(fill="green", alpha=0.7) + 
    labs(title = "Boxplot of proportion white") + 
    scale_x_discrete(name = "Year", 
                     breaks=c(2003, 2004, 2005, 2006, 2007 ,2008, 2009, 2010, 2011), 
                     labels=c("2003", "2004", "2005", "2006", "2007" , "2008", "2009", "2010", "2011")) +
    scale_y_continuous(name = "Proportion of population that identify as white") + 
    theme(text=element_text(size=16), plot.title=element_text(size=18, face="bold"))

```


```{r merge data}
#compute the average sunlight for each year
by_state <- group_by(data, fips)
averageSun <- summarize(by_state, sunDailyAvg = mean(sunlight_daily_avg, na.rm=T))

#merge sunlight data with data_us
averageSun.data_us <- left_join(data_us, averageSun, by = "fips")

#color definition for sunlight
variable <- averageSun.data_us$sunDailyAvg # Ensure we use the merged data column
colors <- colorNumeric(palette = "YlOrBr", domain = variable, reverse = FALSE)

# create the Leaflet map
leaflet(data = averageSun.data_us) |>
  addTiles() |>
  addPolygons(
    fillColor = ~colors(sunDailyAvg),  # Use the correct column
    color = "black",  
    fillOpacity = 0.7,
    weight = 1,
    smoothFactor = 0.5,
    opacity = 1.0) |>
  addLegend(
    pal = colors,
    values = variable,
    opacity = 1,
    title = "Daily Sunlight Average (W/m²)") |>
  addScaleBar(position = "bottomleft")

```

```{r compute spatial correlations}


#### Fit a simple Poisson log-linear model
model1 <- glm(formula=cancer_count~ sunlight_daily_avg + proportion_white, family="poisson", data=data)
round(cbind(model1$coefficients, confint(model1)),4)


#### compute the residuals from this model
data$residuals <- residuals(model1)
residuals2011 <- filter(data, year==2011)
residuals2011_us_data <- left_join(data_us, residuals2011, by = "fips")



#### Construct the spatial objects

# Generate neighbors list
W.nb <- poly2nb(residuals2011_us_data, row.names = residuals2011_us_data$fips)

# Convert to a spatial weights matrix
W <- nb2mat(W.nb, style = "B")

# Convert to a list-based weights object
W.list <- nb2listw(W.nb, style = "B")


#### Conduct Moran's I test
moran.mc(x = residuals2011_us_data$residuals, listw = W.list, nsim = 10000)



```

```{r mcmc chains}


lookup <- data.frame(fips = residuals2011_us_data$fips, spatialorder = 1:nrow(residuals2011_us_data))
data.temp <- merge(x = data, y = lookup, by = "fips")
data.ordered <- arrange(data.temp, year, spatialorder)

#Uncomment to rerun chains if needed

# chain1 <- ST.CARar(formula=cancer_count~ sunlight_daily_avg + proportion_white, family="poisson", data=data.ordered, W=W, burnin=200000, n.sample=2200000, thin=1000, verbose=FALSE, AR = 1)
# chain2 <- ST.CARar(formula=cancer_count~ sunlight_daily_avg + proportion_white, family="poisson", data=data.ordered, W=W, burnin=200000, n.sample=2200000, thin=1000, verbose=FALSE, AR = 1)
# chain3 <- ST.CARar(formula=cancer_count~ sunlight_daily_avg + proportion_white, family="poisson", data=data.ordered, W=W, burnin=200000, n.sample=2200000, thin=1000, verbose=FALSE, AR = 1)
# 
# 
# saveRDS(chain1, file = here("analysis", "chain1.rds"))
# saveRDS(chain2, file = here("analysis", "chain2.rds"))
# saveRDS(chain3, file = here("analysis", "chain3.rds"))

```


```{r convergence}
set.seed(1)
chain1 <- readRDS(file = here("analysis", "chain1.rds"))
chain2 <- readRDS(file = here("analysis", "chain2.rds"))
chain3 <- readRDS(file = here("analysis", "chain3.rds"))

#### Check convergence - traceplot
library(coda)
beta.samples <- mcmc.list(chain1$samples$beta, chain2$samples$beta, chain3$samples$beta)
plot(beta.samples)


#### Check convergence - Gelman-Rubin plot
gelman.diag(beta.samples)


#### Model summary
print(chain1)
print(chain2)
print(chain3)
```
Looking at the trace plots, we can see that we have roughly consistent variation and no trend from the first to final samples in each plot. Also, based on the gelman rubin statistics, we see our two covariates "proportion white" and "sunlight daily average" have point estimates and upper CI of less than 1.1 suggesting convergence.

To then determine if sunlight or proportion white have a significant affect on skin cancer, we can look and see if our uncertainty intervals (2.5%-97.5%) encapsulate zero. We can see that both variables have zero within their intervals, suggesting that both variables have a significant influence on skin cancer.Rho.S is low suggesting low spatial correlation, while Rho.t is close to one, suggesting that there is a temporal correlation. These findings are reflected across all three chains.


## Inference

```{r inference}
set.seed(1)

##########################
#### Section 5 - Inference
##########################
#### Effects of covariates on disease risk
sd(data.ordered$proportion_white)
sd(data.ordered$sunlight_daily_avg) #some insane value...
beta.samples.combined <- rbind(chain1$samples$beta, chain2$samples$beta, chain3$samples$beta)
round(quantile(exp(sd(data.ordered$cancer_count) * beta.samples.combined[ ,2]), c(0.5, 0.025, 0.975)),3)
round(quantile(exp(sd(data.ordered$sunlight_daily_avg) * beta.samples.combined[ ,3]), c(0.5, 0.025, 0.975)),3)
```

ANSWERS Q1) Here we can see that if the proportion of white people increases by 0.112, the cancer count will increase by 14.9%. We can also see that if the daily sunlight average increases by 1630.702 W/m^2, then the cancer count does not increase. However since the interval includes 1, daily sunlight average doesnt have a significant affect, while proportion of white people does include zero, suggesting a significant affect.

```{r inference_risk_trends}
set.seed(1)

#### Compute the risk distributions
fitted.samples.combined <- rbind(chain1$samples$fitted, chain2$samples$fitted, chain3$samples$fitted)
n.samples <- nrow(fitted.samples.combined)
n.all <- ncol(fitted.samples.combined)
risk.samples.combined <- fitted.samples.combined / matrix(rep(data.ordered$cancer_count, n.samples), nrow=n.samples, ncol=n.all, byrow=TRUE) 

#### Compute the areal unit average risk for each year
N <- length(table(data.ordered$year))
risk.trends <- array(NA, c(n.samples, N))
for(i in 1:n.samples)
{
    risk.trends[i, ] <- tapply(risk.samples.combined[i, ], data.ordered$year, mean)
}


#### Plot the average risk trends
time.trends <- as.data.frame(t(apply(risk.trends, 2, quantile, c(0.5, 0.025, 0.975))))
time.trends <- time.trends %>% mutate(year=names(table(data.ordered$year)))
colnames(time.trends)[1:3] <- c("Median","LCI", "UCI")

ggplot(time.trends, aes(x = factor(year), y = Median, group=1)) +
    geom_line(col="red") + 
    geom_line(aes(x=factor(year), y=LCI), col="red", lty=2) +
    geom_line(aes(x=factor(year), y=UCI), col="red", lty=2) + 
    scale_x_discrete(name = "Year", 
                     breaks=c(2003, 2004, 2005, 2006, 2007 ,2008, 2009, 2010, 2011), 
                     labels=c("2003", "2004", "2005", "2006", "2007" , "2008", "2009", "2010", "2011")) +    
    scale_y_continuous(name = "Risk") + 
    theme(text=element_text(size=16), plot.title=element_text(size=18, face="bold")) 
```
THIS ANSWERS TEMPORAL TREND IN RISK PART OF Q2) This is the posterior distribution of average risk per year, with 95% CI intervals. While the intervals are large (questionably so), the trend is consistent, with significant chance from 2003-2004, then staing relatively consistant from 2003 onwards

```{r inference_map_preparation}
set.seed(1)

#### Spatial pattern in disease risk in the last year - mean and PEP
risk.samples.2011 <- risk.samples.combined[ ,data.ordered$year==2011]
risk.2011 <- apply(risk.samples.2011, 2, median)
pep.2011 <- apply(risk.samples.2011 > 1, 2, mean)
#### Map the results
residuals2011_us_data$risk.2011 <- risk.2011
residuals2011_us_data$pep.2011 <- pep.2011
# residuals2010.LA.ll <- spTransform(residuals2010.LA, CRS("+proj=longlat +datum=WGS84 +no_defs"))
residuals2011_us_data <- st_transform(residuals2011_us_data, crs="+proj=longlat +datum=WGS84 +no_defs")
```

```{r inference_map_risk}
set.seed(1)

# colours <- colorNumeric(palette = "YlOrBr", domain = residuals2010.LA.ll@data$risk.2010, reverse=FALSE)
colours <- colorNumeric(palette = "YlOrBr", domain = residuals2011_us_data$risk.2011, reverse=FALSE)
leaflet(data=residuals2011_us_data) %>% 
    addTiles() %>% 
    addPolygons(fillColor = ~colours(risk.2011), 
                color="",
                fillOpacity = 0.7, weight = 1, smoothFactor = 0.5,
                opacity = 1.0) %>%
    addLegend(pal = colours, values = risk.2011, 
              opacity = 1, title="Risk") %>%
    addScaleBar(position="bottomleft")
```

```{r inference_map_pep}
set.seed(1)
# colours <- colorNumeric(palette = "YlOrBr", domain = residuals2010.LA.ll@data$pep.2010, reverse=FALSE)
colours <- colorNumeric(palette = "YlOrBr", domain = residuals2011_us_data$pep.2011, reverse=FALSE)
leaflet(data=residuals2011_us_data) %>% 
    addTiles() %>% 
    addPolygons(fillColor = ~colours(pep.2011), 
                color="",
                fillOpacity = 0.7, weight = 1, smoothFactor = 0.5,
                opacity = 1.0) %>%
    addLegend(pal = colours, values = pep.2011, 
              opacity = 1, title="PEP") %>%
    addScaleBar(position="bottomleft")
```
THE ABOVE TWO MAPS HELP ANSWER THE SPATIAL DISTRIBUTION OF Q2) we can see that the chance of skin cancer risk increasing by 1 is higher in areas with already high risk while areas with lower risk have a lower chance.
